Replacing standard galaxy profiles with \\ mixtures of Gaussians

### 2012-10-22 documents/mixtures/mixture_models.tex
small spelling fixes

We wish to make an approximation to the two-dimensional circular
exponential (exp) profile $Q^{\exp}(\cdot)$
\begin{eqnarray}\displaystyle
Q^{\exp}(\spos) &\equiv& \exp(-\alpha^{\exp}\,[|\spos| - 1])
\\
\alpha^{\exp} &\equiv& 1.67834699
\quad ,
\end{eqnarray}
where $\spos$ is a dimensionless focal-plane position, and
$\alpha^{\exp}$ is a dimensionless inverse length set to ensure that
the profile has unit half-light radius.  The position $\spos$ is
dimensionless because it parameterizes the unit-size dimensionless
function.  We seek the best (where ``best'' will be defined below)
$M^{\exp}$-Gaussian MoG (where $M^{\exp}$ is an integer) approximation
\begin{eqnarray}\displaystyle
Q^{\exp}(\spos) &\approx& \sum_{m=1}^{M^{\exp}} a^{\exp}_m\,\normal(\spos\given\zero,\var^{\exp}_m)
\\
\var^{\exp}_m &\equiv& v^{\exp}_m\,\identity
\quad ,
\end{eqnarray}
where all of the means are exactly zero and all of the variances
$\var^{\exp}_m$ in the MoG can be represented as a scalar
$v^{\exp}_m$ multiplied by the identity matrix $\identity$
because we are requiring this dimensionless function to be precisely
circular (so every component is itself circular and concentric).
Similarly for the de~Vaucouleurs (dev) profile
\begin{eqnarray}\displaystyle
Q^{\dev}(\spos) &\equiv& \exp(-\alpha^{\dev}\,[|\spos|^{1/4} - 1])
\\
\alpha^{\dev} &\equiv& 7.66924944
\\
Q^{\dev}(\spos) &\approx& \sum_{m=1}^{M^{\dev}} a^{\dev}_m\,\normal(\spos\given\zero,\var^{\dev}_m)
\\
\var^{\dev}_m &\equiv& v^{\dev}_m\,\identity
\quad .
\end{eqnarray}
The half-light inverse-radius parameters $\alpha^{\exp}$ and
$\alpha^{\dev}$ are from \citet{ciotti}.  The challenge we
meet below is to determine the parameters
\begin{eqnarray}
\{a^{\exp}_m,v^{\exp}_m\}_{m=1}^{M^{\exp}}~,~\{a^{\dev}_m,v^{\dev}_m\}_{m=1}^{M^{\dev}}
\end{eqnarray}
to best approximate the traditional galaxy profile functions, under
some sensible definiton of the word ``best'', as a function of the
model complexity parameters (numbers of components) $(M^{\exp},
M^{\dev})$.

In addition to these, there are general S\'ersic (ser) profiles, of
which the exp and dev profiles are special cases.  The general ser
profile has one parameter (the ``index'') $n$:
\begin{eqnarray}\displaystyle
Q^{\ser}(\spos\given n) &\equiv& \exp(-\alpha^{\ser}(n)\,[|\spos|^{1/n} - 1])
\\
\{\alpha^{\ser}(2), \alpha^{\ser}(3), \alpha^{\ser}(5)\} &\equiv& \{3.67206075, 5.67016119, 9.66871461\}
\quad ,
\end{eqnarray}
where we have given the constant $\alpha^{\ser}(n)$ for just a few
values of $n$ (\citealt{ciotti}; the $\exp$ and $\dev$ profiles give
the $n=1$ and $n=4$ values).

@@

In our work on \project{The Tractor}, we use the $M^{\lux}=6$ lux and
the $M^{\luv}=8$ luv profiles.  Our advice to users would be to do the
same.  We use the lux and luv over the exp and dev partly because of
their better behaviors numerically, and partly because they and we are
both part of the \project{SDSS} tradition.  These---$M^{\lux}=6$ and
$M^{\luv}=8$---are good compromises between mixture complexity ($M$)
and quality of fit (badness).  Also, even the best-fitting late-type
and early-type galaxies deviate from exponential and de~Vaucouleurs
fits by more than do these high-quality MoG approxmations; no
precision is lost.

@@

In the above we said ``pixel-convolved PSF''.  In every context, when
modeling images, it is valuable to use the pixel-convolved PSF.  With
this definition of the PSF, the pixelized image is the PSF-convolved
true model evaluated at the pixel centers.  This operation is fast.
Other definitions for the PSF (the non-pixel-convolved, for example)
require that the user to \emph{two} convolutions, the first withe the
PSF and the second with the square (or worse) pixel.  Our advice: Only
fit and use pixel-convolved PSFs.

@@

One amusing aspect of MoG profiles has to do with projection from
three to two dimensions.  The projection of a three-dimensional
Gaussian is a two-dimensional Gaussian; the two-dimensional, rigid,
circular approximations we have made for the ser profiles can be
deprojected to rigid, spherical approximations to the
three-dimensional profiles trivially.  The two-dimensional models we
have started with are not accurate models of galaxies in detail---no
galaxy follows exactly any ser profile---so deprojection of our
approximations are are not that interesting in themselves.  However,
the general program of fitting two-dimensional sources with MoGs may
have strong implications in the future for three-dimensional modeling
and deprojection.

>>>>

We wish to make an approximation to the two-dimensional circular
exponential (exp) profile $Q^{\exp}(\cdot)$
\begin{eqnarray}\displaystyle
Q^{\exp}(\spos) &\equiv& \exp(-\alpha^{\exp}\,[|\spos| - 1])
\\
\alpha^{\exp} &\equiv& 1.67834699
\quad ,
\end{eqnarray}
where $\spos$ is a dimensionless focal-plane position, and
$\alpha^{\exp}$ is a dimensionless inverse length set to ensure that
the profile has unit half-light radius.  The position $\spos$ is
dimensionless because it parameterizes the unit-size dimensionless
function.  We seek the best (where ``best'' will be defined below)
$M^{\exp}$-Gaussian MoG (where $M^{\exp}$ is an integer) approximation
\begin{eqnarray}\displaystyle
Q^{\exp}(\spos) &\approx& \sum_{m=1}^{M^{\exp}} a^{\exp}_m\,\normal(\spos\given\zero,\var^{\exp}_m)
\\
\var^{\exp}_m &\equiv& v^{\exp}_m\,\identity
\quad ,
\end{eqnarray}
where all of the means are exactly zero and all of the variances
$\var^{\exp}_m$ in the MoG can be represented as a scalar
$v^{\exp}_m$ multiplied by the identity matrix $\identity$
because we are requiring this dimensionless function to be precisely
circular (so every component is itself circular and concentric).
Similarly for the de~Vaucouleurs (dev) profile
\begin{eqnarray}\displaystyle
Q^{\dev}(\spos) &\equiv& \exp(-\alpha^{\dev}\,[|\spos|^{1/4} - 1])
\\
\alpha^{\dev} &\equiv& 7.66924944
\\
Q^{\dev}(\spos) &\approx& \sum_{m=1}^{M^{\dev}} a^{\dev}_m\,\normal(\spos\given\zero,\var^{\dev}_m)
\\
\var^{\dev}_m &\equiv& v^{\dev}_m\,\identity
\quad .
\end{eqnarray}
The half-light inverse-radius parameters $\alpha^{\exp}$ and
$\alpha^{\dev}$ are from \citet{ciotti}.  The challenge we
meet below is to determine the parameters
\begin{eqnarray}
\{a^{\exp}_m,v^{\exp}_m\}_{m=1}^{M^{\exp}}~,~\{a^{\dev}_m,v^{\dev}_m\}_{m=1}^{M^{\dev}}
\end{eqnarray}
to best approximate the traditional galaxy profile functions, under
some sensible definition of the word ``best'', as a function of the
model complexity parameters (numbers of components) $(M^{\exp},
M^{\dev})$.

In addition to these, there are general S\'ersic (``ser'') profiles, of
which the exp and dev profiles are special cases.  The general ser
profile has one parameter (the ``index'') $n$:
\begin{eqnarray}\displaystyle
Q^{\ser}(\spos\given n) &\equiv& \exp(-\alpha^{\ser}(n)\,[|\spos|^{1/n} - 1])
\\
\{\alpha^{\ser}(2), \alpha^{\ser}(3), \alpha^{\ser}(5)\} &\equiv& \{3.67206075, 5.67016119, 9.66871461\}
\quad ,
\end{eqnarray}
where we have given the constant $\alpha^{\ser}(n)$ for just a few
values of $n$ (\citealt{ciotti}; the $\exp$ and $\dev$ profiles give
the $n=1$ and $n=4$ values).

@@

In our work on \project{The Tractor}, we use the $M^{\lux}=6$ lux and
the $M^{\luv}=8$ luv profiles.  Our advice to users would be to do the
same.  We use the lux and luv over the exp and dev partly because of
their better behaviors numerically, and partly because they and we are
both part of the \project{SDSS} tradition.  These---$M^{\lux}=6$ and
$M^{\luv}=8$---are good compromises between mixture complexity ($M$)
and quality of fit (badness).  Also, even the best-fitting late-type
and early-type galaxies deviate from exponential and de~Vaucouleurs
fits by more than do these high-quality MoG approximations; no
precision is lost.

@@

In the above we said ``pixel-convolved PSF''.  In every context, when
modeling images, it is valuable to use the pixel-convolved PSF.  With
this definition of the PSF, the pixelized image is the PSF-convolved
true model evaluated at the pixel centers.  This operation is fast.
Other definitions for the PSF (the non-pixel-convolved, for example)
require that the user do \emph{two} convolutions, the first with the
PSF and the second with the square (or worse) pixel.  Our advice: Only
fit and use pixel-convolved PSFs.

@@

One amusing aspect of MoG profiles has to do with projection from
three to two dimensions.  The projection of a three-dimensional
Gaussian is a two-dimensional Gaussian; the two-dimensional, rigid,
circular approximations we have made for the ser profiles can be
deprojected to rigid, spherical approximations to the
three-dimensional profiles trivially.  The two-dimensional models we
have started with are not accurate models of galaxies in detail---no
galaxy follows exactly any ser profile---so deprojection of our
approximations are not that interesting in themselves.  However,
the general program of fitting two-dimensional sources with MoGs may
have strong implications in the future for three-dimensional modeling
and deprojection.
